---
title: "HarvardX PH125.9x Data Science: Capstone"
author: "Lê Công Bình"
date: "`r format(Sys.Date())`"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction
This report is part of the capstone project of HarvardX’s Data Science Professional Certificate program. The most of content bases on the knowledge of module 6.2 - Recommendation Systems in HarvardX PH125.8x - Data Science: Machine Learning course. It also presents my skill in "Data Science: Wrangling" and "Data Science: Visualization" courses.

Because of large dataset I can not use almost of algorithm (lm, lgm, knn, or tree base) to forecast. I build model by estimating important factor of dataset.

## 2. EDA and Modeling
I first set up the working environment by installing essential packages. Almost of package already have in assessment. In addition I used lubridate package for date processing

```{r loading-libs, message=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
```

### 2.1. Data Wrangling
Information of the MovieLens 10M dataset can be found: https://grouplens.org/datasets/movielens/10m/

The dataset can be downloaded here: http://files.grouplens.org/datasets/movielens/ml-10m.zip

I use the following code to download the dataset and combine the information of the ratings and the movies into movielens:

```{r, message=FALSE}
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
#                                            title = as.character(title),
#                                            genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

### 2.2 Exploratory Data Analysis
Data size and structure:

- There are 9000055 observes with 6 features for modeling (edx dataset)
each of row is a rate by a user for a movie at specific time
- Exclude title column all other columns will be use to predict

```{r }
edx %>% as_tibble()
```

- We have 68878 users who rated for 10677 movies

```{r }
edx %>% 
  summarize(n_users = n_distinct(userId),
            n_movies = n_distinct(movieId))
```

- There are some movie get many rate from user but other don't

```{r }
edx %>%
  group_by(movieId) %>% summarize(n=n()) %>% 
  arrange(desc(n)) %>% top_n(100) %>%
  ggplot(aes(x=n)) +
  geom_histogram()
```

- Some users are more active than others at rating movies

```{r }
edx %>%
  group_by(userId) %>% summarize(n=n()) %>% arrange(desc(n)) %>% 
  top_n(100) %>%
  ggplot(aes(x=n)) +
  geom_histogram()
```

- The most of rating is 3 or 4 star. We can see a few lower rating than higher rating

```{r }
edx %>% 
    ggplot(aes(x=rating)) +
    geom_histogram()

```

- Timestamp data: There is some evidence of a time effect on average rating.

```{r }
edx <- mutate(edx, date = as_datetime(timestamp),week = round_date(date,"week"))
edx <- mutate(edx, date = as_datetime(timestamp),week = round_date(date,"week"))
```

- Strong evidence of a genre effect.

```{r }
# genres_avg <- edx %>% separate_rows(genres,sep = "\\|") %>% group_by(genres) %>% summarize(avg_rate=mean(rating))
edx %>% group_by(genres) %>%
  summarize(n = n(), avg = mean(rating), se = sd(rating)/sqrt(n())) %>%
	filter(n >= 1000) %>% 
	mutate(genres = reorder(genres, avg)) %>%
	ggplot(aes(x = genres, y = avg, ymin = avg - 2*se, ymax = avg + 2*se)) + 
	geom_point() +
	geom_errorbar() + 
	theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

- Overall Quality: There isn't any NA value in dataset. 

```{r }
c(sum(is.na(edx$UserId)),sum(is.na(edx$movieId)),sum(is.na(edx$rating)))

```

### 2.3 Loss Function
- Loss function based on the residual mean squared error (RMSE) on a test set

```{r }
RMSE <- function(true_ratings, predicted_ratings){
    sqrt(mean((true_ratings - predicted_ratings)^2))
  }

```


### 2.4. Modeling

Base on EDA, the modeling will estimate by train rating, user, movie, genre and timestamp

```{r evaluation}
# Naive by average
mu <- mean(edx$rating) 
naive_rmse <- RMSE(edx$rating, mu)
rmse_results <- tibble(method = "Just the average", RMSE = naive_rmse)

# Add movie effect
movie_avgs <- edx %>% group_by(movieId) %>% summarize(b_i = mean(rating - mu))
predicted_ratings <- mu + validation %>% left_join(movie_avgs, by='movieId') %>% .$b_i
model_1_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results, tibble(method="Movie Effect Model", RMSE = model_1_rmse ))

# Add user effect
user_avgs <- edx %>% left_join(movie_avgs, by='movieId') %>% 
    group_by(userId) %>% 
    summarize(b_u = mean(rating - mu - b_i))
predicted_ratings <- validation %>% 
    left_join(movie_avgs, by='movieId') %>% 
    left_join(user_avgs, by='userId') %>%
    mutate(pred = mu + b_i + b_u) %>% .$pred
model_2_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results, tibble(method="Movie + User Effects Model", RMSE = model_2_rmse ))

# Add genre effect
genre_avgs <- edx %>% 
    left_join(movie_avgs, by='movieId') %>% 
    left_join(user_avgs, by='userId') %>% 
    group_by(genres) %>% 
    summarize(b_g = mean(rating - mu - b_i - b_u))
predicted_ratings <- validation %>% 
    left_join(movie_avgs, by='movieId') %>% 
    left_join(user_avgs, by='userId') %>%
    left_join(genre_avgs, by='genres') %>%
    mutate(pred = mu + b_i + b_u + b_g) %>% .$pred
model_3_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results, tibble(method="Movie + User + Genre Effects Model", RMSE = model_3_rmse ))

# Add time stamp effect
week_avgs <- edx %>% 
    left_join(movie_avgs, by='movieId') %>% 
    left_join(user_avgs, by='userId') %>% 
    left_join(genre_avgs, by='genres') %>% 
    group_by(week) %>% 
    summarize(b_w = mean(rating - mu - b_i - b_u - b_g))
predicted_ratings <- validation %>% mutate(date = as_datetime(timestamp),week = round_date(date,"week")) %>%
    left_join(movie_avgs, by='movieId') %>% 
    left_join(user_avgs, by='userId') %>%
    left_join(genre_avgs, by='genres') %>%
    left_join(week_avgs, by='week') %>%
    mutate(pred = mu + b_i + b_u + b_g + b_w) %>% .$pred
model_4_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results, tibble(method="Movie + User + Genre + Week Effects Model", RMSE = model_4_rmse ))

#rmse_results
#rmse_results %>% knitr::kable()


rmse_target <- 0.86490
rmse_results %>% mutate(RMSE_target = ifelse(RMSE < rmse_target, TRUE, FALSE ))

```

### 2.5. Model Tuning

#### Create train and test using edx dataset only

```{r traintestsplit}
set.seed(1, sample.kind="Rounding") 
test_index <- createDataPartition(y = edx$rating, times = 1, p = 0.1, list = FALSE)
train <- edx[-test_index,]
test <- edx[test_index,]
```

#### Regularization

```{r regularization}
reglr_fit <- function(lambda, trainset, testset){
    mu <- mean(trainset$rating) 
    b_i <- trainset %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu)/(n()+lambda))
    b_u <- trainset %>% left_join(b_i, by="movieId") %>% 
        group_by(userId) %>% 
        summarize(b_u = sum(rating - b_i - mu)/(n()+lambda))
    b_g <- trainset %>% left_join(b_i, by="movieId") %>% left_join(b_u, by='userId') %>%         
        group_by(genres) %>% 
        summarize(b_g = sum(rating - b_i - b_u - mu)/(n()+lambda))    
    b_w <- trainset %>% left_join(b_i, by="movieId") %>% left_join(b_u, by='userId') %>% left_join(b_g, by='genres') %>% 
        group_by(week) %>% 
        summarize(b_w = sum(rating - b_g - b_i - b_u - mu)/(n()+lambda))
    
    predicted_ratings <- testset %>% 
        left_join(b_i, by = "movieId") %>%
        left_join(b_u, by = "userId") %>%
        left_join(b_g, by = "genres") %>%
        left_join(b_w, by = "week") %>%   
        filter(!is.na(b_i), !is.na(b_u), !is.na(b_g), !is.na(b_w)) %>%
        mutate(pred = mu + b_i + b_u + b_g + b_w) %>%
        select(pred, rating)   
    return(RMSE(predicted_ratings$pred, predicted_ratings$rating))
}

lambdas <- seq(-0.5, 8, 0.1)
rmses <- sapply(lambdas, reglr_fit, trainset = train, testset = test)

tibble(Lambda = lambdas, RMSE = rmses) %>%
  ggplot(aes(x = Lambda, y = RMSE)) +
    geom_point()
```

From chart above I choosed lambda = 5 for evaluation

```{r rerun with tunned parameter }
lambda <- 5

mu <- mean(edx$rating) 
b_i <- edx %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu)/(n()+lambda))
b_u <- edx %>% left_join(b_i, by="movieId") %>% 
    group_by(userId) %>% 
    summarize(b_u = sum(rating - b_i - mu)/(n()+lambda))
b_g <- edx %>% left_join(b_i, by="movieId") %>% left_join(b_u, by='userId') %>%         
    group_by(genres) %>% 
    summarize(b_g = sum(rating - b_i - b_u - mu)/(n()+lambda))    
b_w <- edx %>% left_join(b_i, by="movieId") %>% left_join(b_u, by='userId') %>% left_join(b_g, by='genres') %>% 
    group_by(week) %>% 
    summarize(b_w = sum(rating - b_g - b_i - b_u - mu)/(n()+lambda))

predicted_ratings <- validation %>% mutate(date = as_datetime(timestamp),week = round_date(date,"week")) %>%
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    left_join(b_g, by = "genres") %>%
    left_join(b_w, by = "week") %>%   
    #filter(!is.na(b_i), !is.na(b_u), !is.na(b_g), !is.na(b_w)) %>%
    mutate(pred = mu + b_i + b_u + b_g + b_w) %>% .$pred
model_5_rmse <- RMSE(predicted_ratings, validation$rating)

rmse_results <- bind_rows(rmse_results, tibble(method="Movie + User + Genre + Week + Regularization Effects Model", RMSE = model_5_rmse ))
rmse_target <- 0.86490
rmse_results %>% mutate(RMSE_target_passing = ifelse(RMSE < rmse_target, TRUE, FALSE ))

```

It sounds better and passing target of project.

#### Linear model

Can not done due to perfomance

```{r}
#fit <- lm(rating ~ as.factor(movieId) + as.factor(userId), data = edx)
#y_hat <- predict(fit, newdata = test) 
#RMSE(test$rating, y_hat)
```

I tried with glm but it don't have better score: RMSE for glm is 1.06

```{r}
#fit <- train(rating ~ movieId + userId, method = "glm", data = train)
#glm_preds <- predict(fit, test)
#RMSE(test$rating,glm_preds)

#glm_vals <- predict(fit, validation)
#RMSE(validation$rating,glm_vals)
```

## 3. Result
- The best of model is use all feature: Movie, User, Genre, Week (round up date) and Regularization with lambda = 5: RMSE = 0.8643062. 
- It's a bit better than project target: 0.86490

```{r final-result }
rmse_target <- 0.86490
rmse_results %>% mutate(RMSE_target_passing = ifelse(RMSE < rmse_target, TRUE, FALSE ))
```

## 4. Conclusion 

Some machine learning algorithms are computationally expensive. The required amount of memory far exceeded the available in a commodity laptop, even with increased virtual memory. The model works only for existing users, movies and rating values, so the algorithm must run every time a new user or movie is included, or when the rating changes.

In the future I will try with Matrix factorization or use open source package recosystem to evaluation

## 5. References
1. Rafael A. Irizarry (2021), Introduction to Data Science: https://rafalab.github.io/dsbook/
2. Model Fitting and Recommendation Systems Overview: https://learning.edx.org/course/course-v1:HarvardX+PH125.8x+2T2021/block-v1:HarvardX+PH125.8x+2T2021+type@sequential+block@568d02a4412440489621921b87e7536f/block-v1:HarvardX+PH125.8x+2T2021+type@vertical+block@d7b4b0783dd443d1bd14504fe2333c24

```{r}
